\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing inequality with tax data - Switzerland from 1945 to 2010}

### 2014-01-07 paper/subdocuments/data_and_methods.tex
Übertrag aus Skizze

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Journal Article
% LaTeX Template
% Version 1.3 (9/9/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%  PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[twoside]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

\usepackage{lipsum} % Package to generate dummy text throughout this template

\usepackage[sc]{mathpazo} % Use the Palatino font
%\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[utf8]{inputenc}
\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text
\usepackage{paralist} % Used for the compactitem environment which makes bullet points with less space between them

\usepackage{color} % used to mark parts that need to be edited

\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

\usepackage{titlesec} % Allows customization of titles
\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Running title $\bullet$ November 2012 $\bullet$ Vol. XXI, No. 1} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
\bibliographystyle{plain}

%%%-------------------------------------------------%%%
%%% Preferences for Knitr %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document global preferences for Knitr %%%
%%%-------------------------------------------------%%%

%----------------------------------------------------------------------------------------
%  TITLE SECTION
%----------------------------------------------------------------------------------------

\title{\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing inequality with tax data - Switzerland from 1945 to 2010}} % Article title

\author{
\large
\textsc{Oliver Hümbelin}\\[2mm] % Your name
\normalsize Bern University of Applied Sciences \\ % Your institution
\normalsize \href{mailto:oliver.huembelin@bfh.ch}{oliver.huembelin@bfh.ch} % Your email address
\vspace{5mm}\\
\large
\textsc{Rudolf Farys}\\[2mm] % Your name
\normalsize University of Bern \\ % Your institution
\normalsize \href{mailto:rudolf.farys@soz.unibe.ch}{rudolf.farys@soz.unibe.ch} % Your email address
\vspace{-5mm}
}
\date{}

%----------------------------------------------------------------------------------------
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}

\maketitle

\thispagestyle{fancy} % All pages have headers and footers

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

%%%-------------------------------------------------%%%
%%% Include abstract %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document abstract %%%
%%%-------------------------------------------------%%%

\begin{abstract}

There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (OECD, 2011; Gornick and Jäntti, 2013). In Switzerland, however, the development is unclear, as there is only little systematic evidence that would allow for long-term comparison. Nevertheless income inequality has been a prominent theme in the public discussion in recent years, e.g. the most recent referendums "Abzockerinitiative'' and "1:12-Initiative''. We adress the gap by presenting a long and consistent time series of inequality measures for income and wealth for Switzerland (1943-2010) calculated from federal tax data. We describe the benefits and shortcomings of tax data compared to other data sources and present strategies to handle tax data specific methodological difficulties. In the end we integrate the case of Switzerland into the international picture of inequality development showing parallels and deviations.

\end{abstract}

%----------------------------------------------------------------------------------------
%	ARTICLE CONTENTS
%----------------------------------------------------------------------------------------

\begin{multicols}{2} % Two-column layout throughout the main article text

%%%-------------------------------------------------%%%
%%% Include introduction %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document introduction %%%
%%%-------------------------------------------------%%%

\section{Introduction}

% Ich würde hier ohne Unterkapitel arbeiten und die Einführung einfach durch Abschnitte gliedern.

% 1.Abschnitt: Bedeutung von Ungleichheit von Soziologischer Sicht

\emph{Bedeutung von ökonomischer Ungleicheit} \\

% 2.Abschnitt Trends in inequality research
\emph{Studien zur allgemeinen Entwicklung der Ungleichheit nach unterschiedliche Paradigmen}(Armut, Reiche, Mittelschicht)  [Concerns about the hollowing of the middle class is a argument why the tax studies should be exploited to gain knowledge about the whole distribution of economic resources. Until these days tax data is only used to assess the change in top income shares.] \\
\begin{itemize}
  \item Development of inequality with LIS Data (Gornick and Jäntti 2013)
  \item ranking of inequality across high-income countries (Atkinson, Rainwater and Smeeding (1995)
   \item inequality and low income/relative poverty (Ferreira and Ravallion, 2009; Nolan and Marx, 2009; Smeeding, O’Higgins and Rainwatter, 1990; Rainwater and Smeeding 2003; Förster and Vlemickx, 2004)
  \item inequality and top income (Atkinson and Piketty, 2007; Leigh, 2009)
  \item inequality and the middle class (Estache and Leipziger, 2009; Littrell et al., 2001; Birdsall, 2010; Ravallion, 2010; Frank,2007)
\end{itemize}

% 3.Abschnitt: Studien mit Steuerdaten

\emph{Studien mit Steuerdaten}
\begin{itemize}
\item Old studies with tax data (Vilfredo Pareto, 1895, 1896; Simon Kuznets, 1953)
\item More recently revival of tax-studys with focus on top income (Piketty and Saez, 2003; Atkinson and Piketty, 2007,2010).
\end{itemize}

% 4.Abschnitt: What do we know about Switzerland?

\emph{Official Data Collection} \\

\begin{itemize}
\item  LIS (1982-2004, Survey Data from Swiss Federal Statistical Office three different surveys (recent years from “income and Consumption Survey, main result; decreasing inequality)
\item  OECD (2000 und 2004, Income and consumption survey, 2008, EU-SILC. The change of survey in 2008 is consdered as a strict break. Comparison over time not possible (OECD 2012:315)
\item  Worldbank: Nothing
\item  Federal Office of statistics (1998-2011, HABE, survey data, very detailed measurement, main result; inequality in post transfer income is constant)
\item  World income inequality Database (WIID) is more or less the same data as it is published in the LIS data
\end{itemize}
\emph{Fazit}:LIS longest time series, comparability is questionable, more recent years time series from Federal Office of statistics. Both studies use survey data, which, as we will discuss later, have certain drawbacks.\\

\emph{Studies on Switzerland} \\
Vgl. Forschungsantrag (kürzen) plus neuere Studien, Verteilungsbericht 2011 und 2012 (SGB), Avenir Suisse (2013) und Gorgas and Schaltenegger (2011) und Foellmi and Martinez (2013). \\

% Should we discus the non-scientic literature?

% What is the purpose of our paper?
The purpose of the following paper is threefold. First of all, information about economic inequality in Switzerland is sparse. We  fill this gap with official data from the Swiss Federal Tax Administration (FTA). Second, we assess the suitability of the FTA data to report inequality and its changes over time. Third, we integrate the development of swiss inequality into the international picture. We can show similarities and differences to the worldwide trends and therefore contribute to the classical theories (\cite{kuznets1955economic}, \cite{nielsen_kuznets_1997}).

%%%-------------------------------------------------%%%
%%% How did inequality change over time %%%
%%%-------------------------------------------------%%%

%<<subdoc_content_introduction, child='subdocuments/change.Rnw', eval=T>>=
%@

%%%-------------------------------------------------%%%
%%% Include data and methods %%%
%%%-------------------------------------------------%%%

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial “income” see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the “top-income” literature, albeit it is a crucial issue in “theorized” concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

%Beschreiben was die Daten beinhalten und wo es Probleme gibt
\begin{itemize}
\item	Datengrundlage: 1947/48 bis 1981/1982 Eidg. Wehrsteuer. 1983/1984 bis 2010 direkte Bundessteuer. Zugänglich über estv.admin.ch
	\item	Special feature Switch from bi-annual taxation to standard annual tax-system (1995/1996 bis 2003)
	\item	Tabulation by size of income and statistical measures from individual tax records (Brülhart-Daten)
	\item	Income after deductions-> taxable income (employment income, business income and capital income). Is income definition stable over time? Yes it should. Realized capital gains are excluded from the definition. It includes income from  employment, self-employment, capital income and taxable transfer payments. Plus Eigenmietwert, 
	\item	Reported on national and cantonal level
\end{itemize}

@@

\textbf{Incomplete coverage of the population (left censored data.)} What can be done about the not-taxed? Del et al. (2007) impute for non-fillers the 20 percentage of the annual average income. This flattens the distribution on the left side, which is not a problem if you are interested in the top income shares, but it would surly affect overall measures of inequality. Furthermore Del et al. calculate the proportion of non-fillers by estimating the total of tax units out of the population records. \\

\textbf{changes in taxation system  (switch from annual to biannual taxation)} In the mid-1990s a fundamental change in the Swiss tax system took place by switching form the two-years based praenumerando taxation to the one-year based postnumerando taxation. This change was enacted with a transitional period of several years, during which each canton could choose when to adopt the new system.  This is why during the transitional period from 1995 to 2003 there is no uniform tax data published on the Swiss level but only data on the cantonal level  (Foellmi and Martinez:8f). \\
%es wird erwartet, dass der Wechsel Ungleichheitsmasse beeinflussen. Yearly fluctuations are dampened, when income is measured on a two-yearly basis.

\textbf{Estimating percentiles from bracket income tabulation} Pareto interpolation \\

\textbf{Missing of mean-tested benefits as part of the income} -> imputation with recommendation for minimum level for basic needs defined by the SKOS.\\
%is never mentioned as a problem, but it seems to me a better way to approach the non-taxed issue, than dell way (20 % of average income)

\textbf{deductions} Del et al. (2007:477):” we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates.”
Gorgas and Schaltenegger (2011:5): “”..., information on […] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a “perverse redistribution” effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

% Income share specific problems
% Total income denominator Exogenous Approach -> net income reported in the national accounts. Endogenous Approach -> Dell et al. technic-> imputing 20\% of average personal income to non-fillers (which are mainly persons with low or no income). Honestly , is this appropriate? (Everyone did it)
% Total of tax units in the country. exogenous approach -> construct number of total tax units artificially from other data sources
% Endogenous approach -> reported in the tax tables

%%%-------------------------------------------------%%%
%%% Abschnitte aus ersten Version des Papers %%%
%%%-------------------------------------------------%%%

@@

%%%-------------------------------------------------%%%
%%% Include results %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document results %%%
%%%-------------------------------------------------%%%

\section{Results}

\textcolor{red}{Gesamtschweizer Grafik (mit imputierten Daten) einmal eine Linie mit Bias, einmal ohne
Kantonsweise Grafiken
Kuznets / U-Turn, Test der Hypothesen mit den final Daten. Der link zwischen rein deskriptiv  und Theorie plus Modell ist aktuell noch ein krasser Drahtseilakt...}

%Switzerland (income and wealth)
%- graph with gini over time> identify periods of change
%- closer look at interesting periods with relative distribution methods-> is change because of down- or upgrading?
%Cantonal level (income and wealth)
%- different development on regional level 
%Switzerland in international comparison
%- Comparison with other data sources- LIS (inequality in Switzerland is decreasing, which is quite special, because the common pattern in western countries shows an increase in inequality

%- Comparison with other countries (similar data)

% Es scheinen mir aktuell etwas viele Ergebnisse zu sein und ich frage mich, ob wir Ergebnisse auf der Kantonsebene ausklammern sollen. Wenn wir drei Ergebniss haben (1) Unterscheidung von unterschiedlichen Perioden der Ungleichheitsentwicklung (2) Beurteilung dieser Perioden hinsichtlich dem Bereich der Veränderungen (Arme, Reiceh) und Entwicklung der Vermögensungleichheit, dann haben wir bereits viel. Falls wir spezifische Erkenntnisse nur aus dem Kantonsvergleich bekommen (Bias durch Nuller o.ä.), dann sollten wir trotzdem damit arbeiten.

%%%-------------------------------------------------%%%
%%% Include discussion %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document for discussion %%%
%%%-------------------------------------------------%%%

\section{Discussion}

% Hier könnte allenfalls der Bogen zur Theorie gespannt werden

%Methodological Main findings 
%(1) Tax data is useful because it’s possible to construct consistent measures over time for income and wealth, which is useful to asses changes over time and conducted studies about impact of structural changes.(2) Concerning accuracy of inequality tax data has advantages and shortcomings. It’s superior to survey data, because the latter is affected by non-responded bias. Tax data includes data about the whole population … [was finden wir wegen den Nullern raus] . or at least it should (tax evasion).(3) The FTA-Tax data has different shortcomings, which cannot be handled like a measure of income and wealth which (a) cannot be corrected for household size and (b) is neither a pre-transfer nor a post-transfer income measure. Therefor it’s comparability with other surveys is harsh as long as the main sources of data on inequality are based on measures, which cannot be replicated with taxable income. This FTA-Tax data problem can be handled with non-aggregated tax-data .

%Main findings for Switzerland (1) Our data suggests that income inequality overall slightly increased in Switzerland and we might distinguish three episodes: 1. Substantial increase in times of economic growth 1950 to 1974 ending with the oil crisis 2. Ups and downs from 1970 to 2000 3. Relatively steep increase in inequality in the early 2000s
%(2) Despite the recent increase, overall income inequality in Switzerland is not very high in international comparison. With respect to inequality in wealth, however, Switzerland takes a leading position in the world.
%(3) While the overall Swiss inequality remained pretty stable over decades, inequality between cantons underwent a bizarre development.

%%%-------------------------------------------------%%%
%%% Include acknowledgements %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document for acknowledgement %%%
%%%-------------------------------------------------%%%

\section{Acknowledgements}

We thank Ben Jann, Robert Fluder and Tobias Fritschi for helpful comments on the article. We also like to thank Stefan Ilic for the prepartion of the data set.

%%%-------------------------------------------------%%%
%%% Include the bibliography %%%
%%%-------------------------------------------------%%%

\end{multicols}

\bibliography{bibliography/bib}

%%%-------------------------------------------------%%%
%%% Include the appendix %%%
%%%-------------------------------------------------%%%

%%%-------------------------------------------------%%%
%%% Sub document for appendix %%%
%%%-------------------------------------------------%%%

\section{Appendix}

\end{document}

>>>>

@@

\textbf{Summary of tax data drawbacks }
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial “income” see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the “top-income” literature, albeit it is a crucial issue in “theorized” concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

@@

@@
